---
title: "IDS investigation worksheet"
author: "by Artery Party: Roderick Yu, Yufei Xiang, Hoddy Zhu, Lincoln Luk"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(tidymodels)
library(knitr)
library(dplyr)



```


```{r load-data}
# load your data 
MArea <- read_csv("data/HDM_CouncilArea.csv")



```

```{r clean-data}
Data<- MArea %>%
  select(Year,CA,AgeGroup,Sex,Diagnosis,CrudeRate,EASR,NumberOfDeaths) %>%
  drop_na()



```

```{r}

Data %>%
  group_by(AgeGroup) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE)) 

Data %>%
  group_by(Sex) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
              median_cruderate = median(CrudeRate , na.rm = TRUE)) 

Data %>%  
  group_by(Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
              median_cruderate = median(CrudeRate , na.rm = TRUE)) 
```






```{r}
#Question - How does the crude rate for heart diseases compare between regions in Scotland?
#Chart of Area

CA_Regions <- Data %>% 
  filter(!CA %in% c("S92000003")) %>%
  mutate(
    Region = case_when(
      CA %in% c("S12000033", "S12000034", "S12000020", "S12000017") ~ "North",
      CA %in% c("S12000038", "S12000039", "S12000015","S12000047") ~ "North East",
      CA %in% c("S12000005", "S12000014", "S12000026", "S12000036", "S12000042", "S12000044","S12000048") ~ "Central",
      CA %in% c("S12000006", "S12000010", "S12000019", "S12000028", "S12000045") ~ "South",
      CA %in% c("S12000008", "S12000009", "S12000011", "S12000018", "S12000021", "S12000023", "S12000027", "S12000030", "S12000035", "S12000040", "S12000041","S12000013","S12000029","S12000049","S12000050") ~ "West",
      TRUE ~ "Others"
      )
  )


CA_summary <- CA_Regions %>%
  group_by(Region) %>%
  summarise(
    mean_deaths = mean(NumberOfDeaths, na.rm = TRUE),
    mean_rate  = mean(CrudeRate, na.rm = TRUE)
  )


  ggplot(CA_summary, aes(x = Region, y = mean_rate)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Crude Mortality Rate by Council Area",
    x = "Region",
    y = "Mean Crude Mortality Rate"
  )



```

```{r}
# Chart of EASR

ggplot(data = Data,
       mapping = aes(x = EASR, 
                     y = CrudeRate, 
                     )) +
  geom_point(size = 1)+
  geom_smooth(method = lm , se = FALSE)+
  labs(title = "EASR with CrudeRate",
       x="EASR",
       y=" CrudeRate",
       fill="Sex")

```

```{r}


# Bar chart of Sex
table_sex<-Data %>% 
  filter(!Sex %in% c("All"))%>%
  group_by(Sex) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))

ggplot(table_sex)+
       geom_bar(mapping= aes (x=Sex,
                     y=mean_cruderate,
                     fill=Sex),
                     stat = "identity")+
  labs(title = "The mean value of cruderate for different Sexs",
       x="Sex",
       y="Mean Cruderate",
       fill="Sex")

#Bar chart of AgeGroup
table_age<-Data %>%
  filter(!AgeGroup %in% c("All"))%>%
  group_by(AgeGroup) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))%>%
  mutate(AgeGroup = fct_relevel(AgeGroup, "All" ,after = 0))

ggplot(table_age)+
       geom_bar(mapping= aes (x=AgeGroup,
                              y=mean_cruderate,
                              fill=AgeGroup),
                              stat = "identity")+
  labs(title = "The mean value of cruderate for different AgeGroups",
       x="Age",
       y="Mean Cruderate",
       fill="AgeGroup")

#line chart of year
summary_diagnosis <- Data%>%
  group_by(Year) %>%
   summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE))

ggplot(summary_diagnosis, aes(x = Year, y = mean_cruderate)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "The mean value of cruderate over Years",
    x = "Year",
    y = "Mean Cruderate"
  ) +
  theme_minimal()

# Bar chart of Diagnosis
table_Diagnosis<-Data %>%
  group_by(Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))

ggplot(table_Diagnosis)+
       geom_bar(mapping= aes (x=Diagnosis,
                     y=mean_cruderate,
                     fill=Diagnosis),
                     stat = "identity")+
  labs(title = "The mean value of cruderate for different Diagnosis",
       x="Diagnosis",
       y="Mean Cruderate",
       fill="Diagnosis")

```



```{r}
###
##Chart of two variables
###

Data<-Data%>% 
  filter(!Sex %in% c("All"))%>%
  filter(!AgeGroup %in% c("All"))

#Table 1 (Agegroup and Diagnosis)
table_1<- Data %>%
  group_by(AgeGroup,Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))%>%
  mutate(AgeGroup = fct_relevel(AgeGroup, "All" ,after = 0))

ggplot(table_1, aes(x = Diagnosis, y = mean_cruderate, fill = AgeGroup)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Mean crude rate by heart-disease type and Agegroup",
    x = "Heart-disease type",
    y = "Mean crude rate",
    fill = "AgeGroup"
  )

#Table 2 (Sex and Diagnosis)
Table_2<-Data %>%
  group_by(Sex,Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))
ggplot(Table_2, aes(x = Diagnosis, y = mean_cruderate, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Mean crude rate by heart-disease type and sex",
    x = "Heart-disease type",
    y = "Mean crude rate",
    fill = "Sex"
  )

#Table 3(Area and Diagnosis)
CA_summary_with_year<- CA_Regions %>%
  group_by(Region,Year) %>%
  summarise(
    mean_cruderate = mean(CrudeRate , na.rm = TRUE)
  )

ggplot(CA_summary_with_year, aes(x = Year, y =mean_cruderate,  fill = Region)) +
  geom_col(position = position_dodge(width = 0.8))+
  labs(
    title = "Mean crude rate by Area over Years",
    x = "Year",
    y = "Mean crude rate"
  ) 
 
  
#Graph of the Year
#Table 4(Sex and year)
  Table_4<-Data %>%
  group_by(Sex,Year) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))
  
ggplot(Table_4, aes(x = Year, y = mean_cruderate, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Mean crude rate by by Sex over Years",
    x = "Year",
    y = "Mean crude rate",
    fill = "Sex"
  )

#Table 5(Agegroup and year)



  Table_5<-Data %>%
    group_by(AgeGroup,Year) %>%
    summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
              median_cruderate = median(CrudeRate , na.rm = TRUE)) 
  ggplot(Table_5, aes(x = Year, y = mean_cruderate, color =AgeGroup , group = AgeGroup)) +
    geom_line(size = 1) +
    geom_point() +
    labs(
      title = "Mean crude rate by AgeGroup over Years",
      x = "Year",
      y = "Mean crude rate"
    ) +
    theme_minimal()
  
  #Table6(Diagnosis and year)
  summary_diagnosis <- Data %>%
  group_by(Year, Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE))

ggplot(summary_diagnosis,aes(x = Year, y = mean_cruderate, color = Diagnosis, group = Diagnosis)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Mean crude rate by Diagnosis over Years",
    x = "Year",
    y = "Mean crude rate"
  ) +
  theme_minimal()

```




```{r}

  
Data <- Data %>%
  mutate(
    split_AgeGroup = case_when(
      AgeGroup %in% c("75plus years") ~ "75plus years",
      AgeGroup %in% c("0-44 years", "45-64 years", "65-74 years", "under75 years") ~ "under75 years",
      TRUE ~ "Others"
      )
  )


Data$split_AgeGroup<-relevel(factor(Data$split_AgeGroup),ref="under75 years")
Data$Diagnosis<-relevel(factor(Data$Diagnosis),ref="Heart Failure")
Data$Sex<-relevel(factor(Data$Sex),ref="Females")


ht_yc_fit_1 <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log(CrudeRate+1) ~ Year+split_AgeGroup+Sex+Diagnosis,data = Data)
tidy(ht_yc_fit_1)

ht_yc_fit_aug_1 <- augment(ht_yc_fit_1$fit)
ggplot(ht_yc_fit_aug_1, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted CrudeRate", y = "Residuals")

glance(ht_yc_fit_1)$adj.r.squared


ht_yc_fit_2 <- linear_reg() %>%
  set_engine("lm") %>%
  fit(CrudeRate ~ Year,data = Data)
tidy(ht_yc_fit_2)

ht_yc_fit_aug_2 <- augment(ht_yc_fit_2$fit)

ggplot(data=ht_yc_fit_aug_2 ,aes(x=Year,y=CrudeRate))+
  geom_point(alpha=0.25)+
  geom_line(aes(x = Year, y = .fitted), linewidth = 0.75)+
  geom_point(mapping = aes(x = Year,y = .fitted), color = "#E48957")+
  geom_segment(mapping = aes(x = Year, y = CrudeRate,
                             xend = Year, yend = .fitted), color = "#E48957", alpha = 0.4)+
  labs(title = "Year vs CrudeRate",
    x = "Year",
    y = "Cruderate"
  ) 




```

```{r}

CA <- Data %>% 
  filter(!is.na(CrudeRate))%>%
  mutate(rate = CrudeRate / 1000) %>% 
  mutate(rate_group = if_else(rate <= median(rate, na.rm = TRUE), 0, 1)) %>% 
  mutate(rate_group = factor(rate_group))

set.seed(888)
CrudeRate_split <- initial_split(CA, prop = 0.80)

train_data <- training(CrudeRate_split)
test_data  <- testing(CrudeRate_split)

CrudeRate_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(rate_group ~ AgeGroup + Sex + Diagnosis,
      data = train_data,
      fimily = "binomial")

CrudeRate_pred <- predict(CrudeRate_fit, test_data, type = "prob") %>%
  bind_cols(test_data)

CrudeRate_pred %>%
  roc_curve(
    truth = rate_group,
    .pred_1,
    event_level = "second"
  ) %>%
  autoplot()


CrudeRate_pred %>%
  roc_auc(
    truth = rate_group,
    .pred_1,
    event_level = "second"
  )

cutoff_prob <- 0.25
CrudeRate_pred %>%
  mutate(
    rate_group      = if_else(rate_group == 1 , "high_CrudeRate", "low_CrudeRate"),
    rate_group_pred = if_else(.pred_1 > cutoff_prob, "labelled high-CrudeRate", "labelled low-CrudeRate")
  ) %>%
  count(rate_group_pred, rate_group) %>%
  pivot_wider(names_from = rate_group, values_from = n) %>%
  kable(col.names = c(" ","high_CrudeRate","low_CrudeRate"))
```
```{r}

```
