---
title: "IDS Project"
author: "by Artery Party: Roderick Yu, Yufei Xiang, Hoddy Zhu, Lincoln Luk"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}

library(tidyverse)
library(tidymodels)
library(knitr)
library(kableExtra)

```


```{r load-data, include=FALSE}

MArea <- read_csv("data/HDM_CouncilArea.csv")
CouncilArea <- read_csv("data/CouncilArea.csv")


Data<- MArea %>%
  select(Year,CA,AgeGroup,Sex,Diagnosis,CrudeRate,EASR,NumberOfDeaths) %>%
  drop_na()


```


## Research Question

Our main research question is to explore the differences in heart disease mortality in Scotland, and to identify how mortality rates vary across different demographic categories, such as gender, geographical area, age group, and year. The goal is to identify any long-term patterns or to determine whether certain groups consistently exhibit higher mortality rates than others. We also aim to provide risk predictions for the population of Scotland through the use of modelling. 

This particular dataset interested our group as heart diseases can affect anyone and is also known as one of the main causes of death globally. We all have a risk of getting heart disease and it was interesting to us to find what group of people are most at risk. 

## Data Overview

We are using the 'Heart Disease Mortality by Council Area' dataset from the Scottish Heart Disease Statistics from the Public Health Scotland NHS website, which we named "MArea". 

This dataset consists of 17820 rows and 14 variables (columns). 

Our focus was on the different demographic and geographic categories, so from the start we cleaned up the data from "MArea" into "Data" where there are 7230 rows and 7 variables (columns):


```{r, echo = FALSE}
overview <- tribble(
  ~Variable,       ~Description,                                  ~Type,
  "Year",          "Year of recorded data",                       "Numeric",
  "CA",            "Council area code",                           "Categorical",
  "AgeGroup",      "Defined age range",                           "Categorical",
  "Sex",           "Male or female",                              "Categorical",
  "Diagnosis",     "Specific heart-disease-related diagnosis",    "Categorical",
  "CrudeRate",     "Raw mortality rate",                          "Numeric",
  "EASR",          "European Age-Standardised Rate",              "Numeric"
)

overview %>%
  kable(head(overview), format = ("html")) %>% 
  kable_styling(full_width = FALSE, position = "left")
```

Additionally to support our research aim, we made additional variables to categorise our data better and so we could analyse and evaluate the data more easily. The additional variables will be mentioned in the findings.

## Exploratory Statistics

We began by categorising our data into subgroups, grouping age groups, sex and diagnosis into smaller groups. We then matched each subgroup with their mean crude rates to see how they differ.

```{r, echo = FALSE}
Data %>%
  group_by(AgeGroup) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE)) 

Data %>%
  group_by(Sex) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
              median_cruderate = median(CrudeRate , na.rm = TRUE)) 

Data %>%  
  group_by(Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
              median_cruderate = median(CrudeRate , na.rm = TRUE)) 
```

These summary statistics show us that age is the most dominant factor in which the we can see the biggest difference in crude rates, both in mean and median. This is also confirmed from comparing the two sections All and under 75 years, where the rates for under 75 years are significantly lower.

The difference between sexes is also quite clear, with men have a noticeably higher rate than women. Across the three diagnoses, coronary heart disease has the highest rates, followed by heart attacks , while heart failure has the lowest rates.


## Data Visualisations

To understand more about this dataset, we tried to find more patterns by making visualisations

### 1. EASR

Firstly, we compared the EASR and crude rate in this dataset. The EASR is the European Age-Sex standardised rate per 100,000 population. It is adjusted for the age distribution using a standard population (100,000) and provides a more valid and unbiased comparison than crude rate.

```{r,echo=FALSE, message= FALSE}
# Chart of EASR

ggplot(data = Data,
       mapping = aes(x = EASR, 
                     y = CrudeRate, 
                     )) +
  geom_point(size = 1)+
  geom_smooth(method = lm , se = FALSE)
```

We used the EASR and plotted it against crude rate to see how reliable and it was and we plotted a line of best fit, which had a gradient of 1. This shows that crude rate is very reliable and we can confidently use it to compare evaluate the data, as their correlation is near perfect. Therefore we used the crude rate instead of EASR for the rest of the investigation.


### 2. Geographical area

To start off, we tried to find if there were differences in mortality rate for each region of Scotland.
We looked at the dataset CouncilArea from CouncilArea.csv and we manually matched each area with the council area codes in MArea (the original dataset) to each region (North, Central, South, etc.)

```{r, echo = FALSE, warning= FALSE}
#Chart of Area

CA_Regions <- Data %>% 
  filter(!CA %in% c("S92000003")) %>%
  mutate(
    Region = case_when(
      CA %in% c("S12000033", "S12000034", "S12000020", "S12000017") ~ "North",
      CA %in% c("S12000038", "S12000039", "S12000015","S12000047") ~ "North East",
      CA %in% c("S12000005", "S12000014", "S12000026", "S12000036", "S12000042", "S12000044","S12000048") ~ "Central",
      CA %in% c("S12000006", "S12000010", "S12000019", "S12000028", "S12000045") ~ "South",
      CA %in% c("S12000008", "S12000009", "S12000011", "S12000018", "S12000021", "S12000023", "S12000027", "S12000030", "S12000035", "S12000040", "S12000041","S12000013","S12000029","S12000049","S12000050") ~ "West",
      TRUE ~ "Others"
      )
  )


CA_summary <- CA_Regions %>%
  group_by(Region) %>%
  summarise(
    mean_deaths = mean(NumberOfDeaths, na.rm = TRUE),
    mean_rate  = mean(CrudeRate, na.rm = TRUE)
  )


  ggplot(CA_summary, aes(x = Region, y = mean_rate)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Crude Mortality Rate by Council Area",
    x = "Region",
    y = "Mean Crude Mortality Rate"
  )
```

We found that there was not much variation across the regions except for the North-east, where rates are relatively around 5% higher than the other regions. We concluded that there is no trend between geographical location and crude rate for heart diseases, and we decided that this is not something we had to keep looking into. This could also mean that the risk factors such as oil intake or physical activity are quite similar throughout each region. So we decided to investigate the other factors in this dataset.


### 3. Sex 

Then, to visualise the how the crude rate differs between sex, we made a bar chart with the mean value of crude rates across all men and women. 

```{r, echo = FALSE, warning= FALSE}
# Bar chart of Sex
table_sex<-Data %>%
  group_by(Sex) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))

ggplot(table_sex)+
       geom_bar(mapping= aes (x=Sex,
                     y=mean_cruderate,
                     fill=Sex),
                     stat = "identity")+
  labs(title = "The mean value of cruderate for different Sexes",
       x="Sex",
       y="Mean Cruderate",
       fill="Sex")
```

We found that men have a significantly higher mean crude rate than women, almost double. This shows that men are more twice as likely to die from heart diseases than women This may be caused by eating habits, physical activity, or even biological reasons. We figured this was worth looking into.

Significant gender differences:
The average CrudeRate of men (47.9) is 72% higher than that of women (27.8), but the median for both is 0, indicating that most individuals have no events, but men have more extreme high-risk cases.
The global median (1.95) falls between men and women, indicating that the dataset may contain mixed gender ratios.


### 4. Age

We then investigated the different age groups differed in crude rate. 

```{r, echo = FALSE, warning= FALSE}
#Bar chart of Age
table_age<-Data %>%
  group_by(AgeGroup) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))%>%
  mutate(AgeGroup = fct_relevel(AgeGroup, "All" ,after = 0))

ggplot(table_age)+
       geom_bar(mapping= aes (x=AgeGroup,
                              y=mean_cruderate,
                              fill=AgeGroup),
                              stat = "identity")+
  labs(title = "The mean value of cruderate for different AgeGroups",
       x="Age",
       y="Mean Cruderate",
       fill="AgeGroup")
```

The clear takeaway form this visualisation is 75 plus year olds have the highest mean crude rate. This tells us that this group of people (aged 75 plus) are most likely to suffer from heart disease.

Howeverm the mean and median of "all ages" are close (72.8 vs 72.3) and relatively lower, but after splitting, it can be seen that using global statistics will ignore high-risk elderly groups.


### 5. Year

We then compared mean crude rate with year, visualising it with a line chart. 

```{r, echo = FALSE, warning= FALSE}
#line chart of year
summary_diagnosis <- Data%>%
  group_by(Year) %>%
   summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE))

ggplot(summary_diagnosis, aes(x = Year, y = mean_cruderate)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "The mean value of cruderate over Years",
    x = "Year",
    y = "Mean Cruderate"
  ) +
  theme_minimal()
```

We found that the mean crude rate falls throughout the years until 2019, where it increases at a faster rate. This does not tell us much about if the year has any correlation with crude rate, but it shows how it has changed over the years and we could use it to compare with other variables. 

### 6. Diagnosis

Finally, for the single variables, we looked at the diagnoses and how crude rate differed between the three. 

``` {r, echo = FALSE, warning= FALSE}
# Bar chart of Diagnosis
table_Diagnosis<-Data %>%
  group_by(Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))

ggplot(table_Diagnosis)+
       geom_bar(mapping= aes (x=Diagnosis,
                     y=mean_cruderate,
                     fill=Diagnosis),
                     stat = "identity")+
  labs(title = "The mean value of cruderate for different Diagnosis",
       x="Diagnosis",
       y="Mean Cruderate",
       fill="Diagnosis")
```

This visualisation does not tell us much, but we can clearly see the distribution of deaths among the three diagnoses, with coronary heart diseases with the highest rate, followed by heart attacks, and heart failure with the lowest rate. 


Coronary heart disease: The mean and median values are close (116.0 vs 112.0), indicating a relatively symmetrical data distribution, stable risk, and generally high risk.
Heart failure: The mean (2.32) is much higher than the median (0), indicating that the vast majority of cases have no events, but there are a small number of individuals with extremely high risk.



### 7. Combining variables 

After investigating each variable individually against the mean crude rate, we figured that each variable did not tell us much on its own. This led us into combining variables together, comparing two variables together with crude rate.

We 
```{r, echo = FALSE, message= FALSE}
#Table 7(Age and Year)
Table_5 <- Data %>%
  group_by(AgeGroup,Year) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE)) 
ggplot(Table_5, aes(x = Year, y = mean_cruderate, color =AgeGroup , group = AgeGroup)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Mean crude rate by AgeGroup over Years",
    x = "Year",
    y = "Mean crude rate"
  ) +
  theme_minimal()
```

We then plotted year against crude rate

```{r, echo = FALSE, message= FALSE}
#Table 8(Diagnosis and year)
summary_diagnosis <- Data %>%
group_by(Year, Diagnosis) %>%
summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE))

ggplot(summary_diagnosis,aes(x = Year, y = mean_cruderate, color = Diagnosis, group = Diagnosis)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Mean crude rate by Diagnosis over Years",
    x = "Year",
    y = "Mean crude rate"
  ) +
  theme_minimal()
```




```{r, echo = FALSE, warning = FALSE, message = FALSE}
#Table 9 (Agegroup and Diagnosis)
table_1<- Data %>%
  group_by(AgeGroup,Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))%>%
  mutate(AgeGroup = fct_relevel(AgeGroup, "All" ,after = 0))

ggplot(table_1, aes(x = Diagnosis, y = mean_cruderate, fill = AgeGroup)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Mean crude rate by heart-disease type and Agegroup",
    x = "Heart-disease type",
    y = "Mean crude rate",
    fill = "AgeGroup"
  )
```





```{r, echo = FALSE, message= FALSE}
#Table 10 (Sex and Diagnosis)
Table_2<-Data %>%
  group_by(Sex,Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))
ggplot(Table_2, aes(x = Diagnosis, y = mean_cruderate, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Mean crude rate by heart-disease type and sex",
    x = "Heart-disease type",
    y = "Mean crude rate",
    fill = "Sex"
  )
```



Going through all these visualisations and the exploratory statistics, we concluded that men aged over 75 are most likely to suffer from heart disease and that year








## Predictive Modelling

### Data Processing

After removing all observations labeled as “ALL” in the Sex and AgeGroup variables, the dataset contains 3068 valid records.

We excluded these categories because they represent aggregated totals and could distort the accuracy of the model by mixing population groups with very different characteristics.
```{r}
Data<-Data%>% 
  filter(!Sex %in% c("All"))%>%
  filter(!AgeGroup %in% c("All"))
```
Grouping

```{r}
Data <- Data %>%
  mutate(
    split_AgeGroup = case_when(
      AgeGroup %in% c("75plus years") ~ "75plus years",
      AgeGroup %in% c("0-44 years", "45-64 years", "65-74 years", "under75 years") ~ "under75 years",
      TRUE ~ "Others"
      )
  )


Data$split_AgeGroup<-relevel(factor(Data$split_AgeGroup),ref="under75 years")
Data$Diagnosis<-relevel(factor(Data$Diagnosis),ref="Heart Failure")
Data$Sex<-relevel(factor(Data$Sex),ref="Females")
```

### Coefficient Interpretation
```{r}
ht_yc_fit_1 <- linear_reg() %>%
  set_engine("lm") %>%
  fit(CrudeRate ~ Year+split_AgeGroup+Sex+Diagnosis ,data = Data)
tidy(ht_yc_fit_1)
```

##### Year:
The coefficient for Year is extremely small, indicating that time has minimal or no linear effect on crude mortality rate. This aligns with the visual scatter plot showing stable mortality levels across the years.

##### AgeGroup:
The 75+ age group has the largest estimated coefficient, making it the strongest predictor of crude mortality rate. This reflects the well-known increased cardiovascular risk among older populations.

##### Sex:
The coefficient for males is positive and significant, showing that male groups have higher mortality rates than females. This confirms that gender is an important demographic factor.

##### Diagnosis:
Both Coronary Heart Disease and Heart Attack categories have noticeably large coefficient estimates.
These diagnosis types are therefore also major contributors to increased mortality rate.


#### Importance Ranking (from least to most influential):
Year < Is Male < Is Heart Attack < Is Coronary Heart Disease < Is Age 75+

### Residual Plot Analysis
```{r}

ht_yc_fit_aug_1 <- augment(ht_yc_fit_1$fit)
ggplot(ht_yc_fit_aug_1, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted CrudeRate", y = "Residuals")


```


The residual plot shows several important patterns:

#### 1. Clustered residuals with a large empty gap

Residuals concentrate on the left and right sides of the predicted scale, with very few observations between predicted values 100–200.
This structure may caused by:
Most groups have low crude mortality rates；Only specific groups (e.g., 75+ or certain diagnoses) have very high mortality.Hence,this creates natural gaps in the distribution, independent of the model

#### 2. Fan-shaped pattern 

The residuals spread out more widely as the predicted value increases, forming a fan shape.

Possible reasons include:
- Missing important predictors (e.g., regional differences, socioeconomic factors)
- Non-linear relationships that the linear model cannot capture
- The limitations of simple transformations
    - Even after trying log(CrudeRate + 1), the distribution did not improve

This suggests that the linearity assumption may not fully hold.
```{r results = "hide"}
ht_yc_fit_2 <- linear_reg() %>%
  set_engine("lm") %>%
  fit(log(CrudeRate+1) ~ Year+split_AgeGroup+Sex+Diagnosis ,data = Data)
tidy(ht_yc_fit_2)
```


```{r results = "hide"}
ht_yc_fit_aug_2 <- augment(ht_yc_fit_2$fit)
ggplot(ht_yc_fit_aug_2, mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "gray", lty = "dashed") +
  labs(x = "Predicted CrudeRate", y = "Residuals")
```

### Adjusted R²

The adjusted R² value of this model indicates that:
The linear regression model does not fit the data very well.
It explains only a modest proportion of the variation in crude mortality rate.
```{r}
glance(ht_yc_fit_1)$adj.r.squared
```

### Scatter Plot: CrudeRate vs Year

To further investigate the negligible effect of time, we plotted CrudeRate against Year.

The scatter plot shows:
- Clear vertical stacking of values
- No visible linear relationship between Year and CrudeRate
- Mortality rates remain stable across years

These patterns confirm that Year is not a meaningful predictor in this dataset.
```{r results = "hide"}
ht_yc_fit_3 <- linear_reg() %>%
  set_engine("lm") %>%
  fit(CrudeRate ~ Year,data = Data)
tidy(ht_yc_fit_3)
```


```{r}
ht_yc_fit_aug_3 <- augment(ht_yc_fit_3$fit)

ggplot(data=ht_yc_fit_aug_3 ,aes(x=Year,y=CrudeRate))+
  geom_point(alpha=0.25)+
  geom_line(aes(x = Year, y = .fitted), linewidth = 0.75)+
  geom_point(mapping = aes(x = Year,y = .fitted), color = "#E48957")+
  geom_segment(mapping = aes(x = Year, y = CrudeRate,
                             xend = Year, yend = .fitted), color = "#E48957", alpha = 0.4)+
  labs(title = "Year vs CrudeRate",
    x = "Year",
    y = "Cruderate"
  ) 

```

### Conclusion

Although the linear regression model has limitations, it successfully identifies three key factors that may influence heart disease mortality:

1. Age group (especially 75+)
2. Sex (male)
3. Diagnosis type (Coronary Heart Disease, Heart Attack)

### Suggestions for Improvement

1. Include additional variables or perform subgroup analysis to improve the model’s explanatory power
(e.g., add regional effects, socioeconomic indicators)

2. Use a larger dataset such as all regions of the UK to test whether the observed patterns remain consistent

3. Consider exploring non-linear models or classification models (e.g., logistic regression) for further insights


## logistic regression model

We developed a binary classification model to predict cardiovascular mortality risk (high-risk vs. low-risk) for the Scottish population. The workflow and findings are summarized below.

 
- Outcome: The original crude mortality rate was expressed per 100 000 population. For easier interpretation we divided it by 100 to obtain a rate per 100 population, then dichotomised it at the median: values above the median were coded 1 (high risk) and values below the median 0 (low risk).  
- Predictors: age, sex and diagnosis.  
- Data split: the dataset was randomly divided into 80 % training and 20 % test sets.

```{r}
CA <- Data %>% 
  mutate(rate = CrudeRate / 1000) %>% 
  mutate(rate_group = if_else(rate <= median(rate, na.rm = TRUE), 0, 1)) %>% 
  mutate(rate_group = factor(rate_group))

set.seed(888)
CrudeRate_split <- initial_split(CA, prop = 0.80)

train_data <- training(CrudeRate_split)
test_data  <- testing(CrudeRate_split)

CrudeRate_fit <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(rate_group ~ AgeGroup + Sex + Diagnosis,
      data = train_data,
      fimily = "binomial")

CrudeRate_pred <- predict(CrudeRate_fit, test_data, type = "prob") %>%
  bind_cols(test_data)

AUC_ROC <- CrudeRate_pred %>%
  roc_curve(
    truth = rate_group,
    .pred_1,
    event_level = "second"
  )
  
AUC_ROC %>% autoplot()
```

 ROC curve  
   - X-axis: false-positive rate (FPR), the proportion of true low-risk individuals incorrectly classified as high-risk.  
   - Y-axis: true-positive rate (TPR), the proportion of true high-risk individuals correctly identified.

   The curve bows toward the upper-left corner, indicating good discriminatory performance.


```{r}
cutoff_prob<-0.25
CrudeRate_pred %>%
  mutate(
    rate_group      = if_else(rate_group == 1 , "high_CrudeRate", "low_CrudeRate"),
    rate_group_pred = if_else(.pred_1 > cutoff_prob, "labelled high-CrudeRate", "labelled low-CrudeRate")
  ) %>%
  count(rate_group_pred, rate_group) %>%
  pivot_wider(names_from = rate_group, values_from = n) %>%
  kable(col.names = c(" ","high_CrudeRate","low_CrudeRate"))
```
- Sensitivity = 30 / ( 30 + 20 ) ≈ 0.60
- Specificity = 534 / ( 534 + 32 ) ≈ 0.943 
```{r}
cutoff_prob<-0.5
CrudeRate_pred %>%
  mutate(
    rate_group      = if_else(rate_group == 1 , "high_CrudeRate", "low_CrudeRate"),
    rate_group_pred = if_else(.pred_1 > cutoff_prob, "labelled high-CrudeRate", "labelled low-CrudeRate")
  ) %>%
  count(rate_group_pred, rate_group) %>%
  pivot_wider(names_from = rate_group, values_from = n) %>%
  kable(col.names = c(" ","high_CrudeRate","low_CrudeRate"))
```
- Sensitivity = 24 / ( 24 + 26 ) ≈ 0.48
- Specificity = 553 / ( 553 + 26 ) ≈ 0.977  
```{r}
cutoff_prob<-0.75
CrudeRate_pred %>%
  mutate(
    rate_group      = if_else(rate_group == 1 , "high_CrudeRate", "low_CrudeRate"),
    rate_group_pred = if_else(.pred_1 > cutoff_prob, "labelled high-CrudeRate", "labelled low-CrudeRate")
  ) %>%
  count(rate_group_pred, rate_group) %>%
  pivot_wider(names_from = rate_group, values_from = n) %>%
  kable(col.names = c(" ","high_CrudeRate","low_CrudeRate"))
```
- Sensitivity = 9 / ( 9 + 41 ) ≈ 0.18
- Specificity = NA

  Among the candidate thresholds {0.25, 0.50, 0.75}, comparing 0.25 with 0.50 shows that raising the cut-off from 0.25 to 0.50 reduces sensitivity by 12 percentage points (from 0.60 to 0.48) while improving specificity by 3.4 percentage points (from 0.943 to 0.977).
  
  Since the priority is not to miss any high-risk individuals, we retain 0.25 as the preferred cut-off.

  At this threshold the model achieves:
  Sensitivity = 0.60
  Specificity = 0.943
  
  Although this choice accepts a slightly higher false-positive rate, it captures 60 % of true high-risk cases, compared with only 48 % at the more conservative threshold. 
  
  Future work will explore ways to raise sensitivity further without materially sacrificing specificity.



```{r}
CrudeRate_pred %>%
  roc_auc(
    truth = rate_group,
    .pred_1,
    event_level = "second"
  )
```
The area under the ROC curve (AUC) is 0.842. A higher AUC indicates better performance, so this value confirms that the model is of good quality.

Overall, using only age, sex and diagnosis, the model provides a promising initial screen for individuals at elevated risk of cardiovascular death and is ready for further development and prospective evaluation.



## Conclusion







## References

Public Health Scotland (NHS Scotland, 2024) -- Scottish Heart Disease Statistics -- Heart Disease Mortality by Council Area
https://www.opendata.nhs.scot/dataset/scottish-heart-disease-statistics/resource/fc7b42f1-4de6-48dd-b640-613fb0aa427d
accessed (17/10/2025)

Public Health Scotland (NHS Scotland, 2020) -- Council Area 2011 - Council Area 2019 
https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/967937c4-8d67-4f39-974f-fd58c4acfda5
accessed (7/11/2025)
