---
title: "IDS Project"
author: "by Artery Party: Roderick Yu, Yufei Xiang, Hoddy Zhu, Lincoln Luk"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}

library(tidyverse)
library(tidymodels)
library(knitr)

```


```{r load-data, include=FALSE}

MArea <- read_csv("data/HDM_CouncilArea.csv")
CouncilArea <- read_csv("data/CouncilArea.csv")


Data<- MArea %>%
  select(Year,CA,AgeGroup,Sex,Diagnosis,CrudeRate,EASR,NumberOfDeaths) %>%
  drop_na()


```


## Research Question

Our main research question is to explore the differences in heart disease mortality in Scotland, and to identify how mortality rates vary across different demographic categories, such as gender, geographical area, age group, and year. The goal is to identify any long-term patterns or to determine whether certain groups consistently exhibit higher mortality rates than others. We also aim to provide risk predictions for the population of Scotland through the use of modelling. 

This particular dataset interested our group as heart diseases can affect anyone and is also known as one of the main causes of death globally. We all have a risk of getting heart disease and it was interesting to us to find what group of people are most at risk. 

## Data Overview

We are using the 'Heart Disease Mortality by Council Area' dataset from the Scottish Heart Disease Statistics from the Public Health Scotland NHS website, which we named "MArea". 

This dataset consists of 17820 rows and 14 variables (columns). 

Our focus was on the different demographic and geographic categories, so from the start we cleaned up the data from "MArea" into "Data" where there are 7230 rows and 7 variables (columns):


```{r}
overview <- tribble(
  ~Variable,    ~Description,                               ~Type,
  "Year",       "Year of recorded data",                    "Numeric",
  "CA",         "Council area code",                        "Categorical",
  "AgeGroup",   "Defined age range",                        "Categorical",
  "Sex",        "Male or female",                           "Categorical",
  "Diagnosis",  "Specific heart-disease-related diagnosis", "Categorical",
  "CrudeRate",  "Raw mortality rate",                        "Numeric",
  "EASR",       "European Age-Standardised Rate",          "Numeric"
)

kable(head(overview), format = ("html"))
```

- Year – year of recorded data
- CA – council area code
- AgeGroup – categorical age group
- Sex – male or female
- Diagnosis – specific heart-disease-related diagnosis
- CrudeRate – raw mortality rate
- EASR – European Age-Standardised Rate, allowing better comparisons across groups


Additionally to support our research aim, we made additional variables to categorise our data better and so we could analyse and evaluate the data more easily. The additional variables will be mentioned in the findings.

## Exploratory Statistics

We began by categorising our data into subgroups, grouping age groups, sex and diagnosis into smaller groups. We then matched each subgroup with their mean crude rates to see how they differ.

```{r, echo = FALSE}
Data %>%
  group_by(AgeGroup) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE)) 

Data %>%
  group_by(Sex) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
              median_cruderate = median(CrudeRate , na.rm = TRUE)) 

Data %>%  
  group_by(Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
              median_cruderate = median(CrudeRate , na.rm = TRUE)) 
```

These summary statistics show us that age is the most dominant factor in which the we can see the biggest difference in crude rates, both in mean and median. This is also confirmed from comparing the two sections All and under 75 years, where the rates for under 75 years are significantly lower.

The difference between sexes is also quite clear, with men have a noticeably higher rate than women. Across the three diagnoses, coronary heart disease has the highest rates, followed by heart attacks , while heart failure has the lowest rates.


## Data Visualisations

To understand more about this dataset, we tried to find more patterns by making visualisations

### 1. EASR

Firstly, we compared the EASR and crude rate in this dataset. The EASR is the European Age-Sex standardised rate per 100,000 population. It is adjusted for the age distribution using a standard population (100,000) and provides a more valid and unbiased comparison than crude rate.

```{r,echo=FALSE, message= FALSE}
# Chart of EASR

ggplot(data = Data,
       mapping = aes(x = EASR, 
                     y = CrudeRate, 
                     )) +
  geom_point(size = 1)+
  geom_smooth(method = lm , se = FALSE)
```

We used the EASR and plotted it against crude rate to see how reliable and it was and we plotted a line of best fit, which had a gradient of 1. This shows that crude rate is very reliable and we can confidently use it to compare evaluate the data, as their correlation is near perfect. Therefore we used the crude rate instead of EASR for the rest of the investigation.


### 2. Geographical area

To start off, we tried to find if there were differences in mortality rate for each region of Scotland.
We looked at the dataset CouncilArea from CouncilArea.csv and we manually matched each area with the council area codes in MArea (the original dataset) to each region (North, Central, South, etc.)

```{r, echo = FALSE, warning= FALSE}
#Chart of Area

CA_Regions <- Data %>% 
  filter(!CA %in% c("S92000003")) %>%
  mutate(
    Region = case_when(
      CA %in% c("S12000033", "S12000034", "S12000020", "S12000017") ~ "North",
      CA %in% c("S12000038", "S12000039", "S12000015","S12000047") ~ "North East",
      CA %in% c("S12000005", "S12000014", "S12000026", "S12000036", "S12000042", "S12000044","S12000048") ~ "Central",
      CA %in% c("S12000006", "S12000010", "S12000019", "S12000028", "S12000045") ~ "South",
      CA %in% c("S12000008", "S12000009", "S12000011", "S12000018", "S12000021", "S12000023", "S12000027", "S12000030", "S12000035", "S12000040", "S12000041","S12000013","S12000029","S12000049","S12000050") ~ "West",
      TRUE ~ "Others"
      )
  )


CA_summary <- CA_Regions %>%
  group_by(Region) %>%
  summarise(
    mean_deaths = mean(NumberOfDeaths, na.rm = TRUE),
    mean_rate  = mean(CrudeRate, na.rm = TRUE)
  )


  ggplot(CA_summary, aes(x = Region, y = mean_rate)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Crude Mortality Rate by Council Area",
    x = "Region",
    y = "Mean Crude Mortality Rate"
  )
```

We found that there was not much variation across the regions except for the North-east, where rates are relatively around 5% higher than the other regions. We concluded that there is no trend between geographical location and crude rate for heart diseases, and we decided that this is not something we had to keep looking into. This could also mean that the risk factors such as oil intake or physical activity are quite similar throughout each region. So we decided to investigate the other factors in this dataset.


### 3. Sex 

Then, to visualise the how the crude rate differs between sex, we made a bar chart with the mean value of crude rates across all men and women. 

```{r, echo = FALSE, warning= FALSE}
# Bar chart of Sex
table_sex<-Data %>%
  group_by(Sex) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))

ggplot(table_sex)+
       geom_bar(mapping= aes (x=Sex,
                     y=mean_cruderate,
                     fill=Sex),
                     stat = "identity")+
  labs(title = "The mean value of cruderate for different Sexes",
       x="Sex",
       y="Mean Cruderate",
       fill="Sex")
```

We found that men have a significantly higher mean crude rate than women, almost double. This shows that men are more twice as likely to die from heart diseases than women This may be caused by eating habits, physical activity, or even biological reasons. We figured this was worth looking into.


### 4. Age

We then investigated the different age groups differed in crude rate. 

```{r, echo = FALSE, warning= FALSE}
#Bar chart of Age
table_age<-Data %>%
  group_by(AgeGroup) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))%>%
  mutate(AgeGroup = fct_relevel(AgeGroup, "All" ,after = 0))

ggplot(table_age)+
       geom_bar(mapping= aes (x=AgeGroup,
                              y=mean_cruderate,
                              fill=AgeGroup),
                              stat = "identity")+
  labs(title = "The mean value of cruderate for different AgeGroups",
       x="Age",
       y="Mean Cruderate",
       fill="AgeGroup")
```

The clear takeaway form this visualisation is 75 plus year olds have the highest mean crude rate. This tells us that this group of people (aged 75 plus) are most likely to suffer from heart disease.

Extreme age dependence:
The average CrudeRate (314.0) of the population aged 75 and above is tens to hundreds of times higher than other age groups, with a median of 63.0, indicating a significant concentration of risk in this group.
The median of the population under 75 years old is 0, but the average is 4.15, indicating that a few high-value individuals have raised the mean and the overall distribution is highly skewed to the right.

Public Health resources could be prioritized for the population aged 75 and above, as they have the highest mean and median rates, while other age groups have a median of 0. 


### 5. Year

We then compared mean crude rate with year, visualising it with a line chart. 

```{r, echo = FALSE, warning= FALSE}
#line chart of year
summary_diagnosis <- Data%>%
  group_by(Year) %>%
   summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE))

ggplot(summary_diagnosis, aes(x = Year, y = mean_cruderate)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "The mean value of cruderate over Years",
    x = "Year",
    y = "Mean Cruderate"
  ) +
  theme_minimal()
```

We found that the mean crude rate falls throughout the years until 2019, where it increases at a faster rate. This does not tell us much about if the year has any correlation with crude rate, but it shows how it has changed over the years and we could use it to compare with other variables. 

### 6. Diagnosis

Finally, for the single variables, we looked at the diagnoses and how crude rate differed between the three. 

``` {r, echo = FALSE, warning= FALSE}
# Bar chart of Diagnosis
table_Diagnosis<-Data %>%
  group_by(Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))

ggplot(table_Diagnosis)+
       geom_bar(mapping= aes (x=Diagnosis,
                     y=mean_cruderate,
                     fill=Diagnosis),
                     stat = "identity")+
  labs(title = "The mean value of cruderate for different Diagnosis",
       x="Diagnosis",
       y="Mean Cruderate",
       fill="Diagnosis")
```

This visualisation does not tell us much, but we can clearly see the distribution of deaths among the three diagnoses, with coronary heart diseases with the highest rate, followed by heart attacks, and heart failure with the lowest rate. 

### 7. Combining variables 

After investigating each variable individually against the mean crude rate, we figured that each variable did not tell us much on its own. This led us to 

```{r, echo = FALSE, message= FALSE}
#Table 7(Age and Year)
Table_5 <- Data %>%
  group_by(AgeGroup,Year) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE)) 
ggplot(Table_5, aes(x = Year, y = mean_cruderate, color =AgeGroup , group = AgeGroup)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Mean crude rate by AgeGroup over Years",
    x = "Year",
    y = "Mean crude rate"
  ) +
  theme_minimal()
```



We then plotted year against crude rate

```{r, echo = FALSE, message= FALSE}
#Table 8(Diagnosis and year)
summary_diagnosis <- Data %>%
group_by(Year, Diagnosis) %>%
summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE))

ggplot(summary_diagnosis,aes(x = Year, y = mean_cruderate, color = Diagnosis, group = Diagnosis)) +
  geom_line(size = 1) +
  geom_point() +
  labs(
    title = "Mean crude rate by Diagnosis over Years",
    x = "Year",
    y = "Mean crude rate"
  ) +
  theme_minimal()
```




```{r, echo = FALSE, message= FALSE}
#Table 9 (Agegroup and Diagnosis)
table_1<- Data %>%
  group_by(AgeGroup,Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))%>%
  mutate(AgeGroup = fct_relevel(AgeGroup, "All" ,after = 0))

ggplot(table_1, aes(x = Diagnosis, y = mean_cruderate, fill = AgeGroup)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Mean crude rate by heart-disease type and Agegroup",
    x = "Heart-disease type",
    y = "Mean crude rate",
    fill = "AgeGroup"
  )
```





```{r, echo = FALSE, message= FALSE}
#Table 10 (Sex and Diagnosis)
Table_2<-Data %>%
  group_by(Sex,Diagnosis) %>%
  summarise(mean_cruderate = mean(CrudeRate , na.rm = TRUE),
            median_cruderate = median(CrudeRate , na.rm = TRUE))
ggplot(Table_2, aes(x = Diagnosis, y = mean_cruderate, fill = Sex)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Mean crude rate by heart-disease type and sex",
    x = "Heart-disease type",
    y = "Mean crude rate",
    fill = "Sex"
  )
```



Going through all these visualisations and the exploratory statistics, we concluded that men aged over 75 are most likely to suffer from heart disease and that year




Full data conceals the truth:
The mean and median of "all ages" are close (72.8 vs 72.3), but after splitting, it can be seen that the age stratification effect is extremely strong, and directly using global statistics will ignore high-risk elderly groups.

Significant gender differences:
The average CrudeRate of men (47.9) is 72% higher than that of women (27.8), but the median for both is 0, indicating that most individuals have no events, but men have more extreme high-risk cases.
The global median (1.95) falls between men and women, indicating that the dataset may contain mixed gender ratios.

Coronary heart disease: The mean and median values are close (116.0 vs 112.0), indicating a relatively symmetrical data distribution, stable risk, and generally high risk.
Heart failure: The mean (2.32) is much higher than the median (0), indicating that the vast majority of cases have no events, but there are a small number of individuals with extremely high risk.

Coronary heart disease>heart attack>heart failure, which conforms to the common sense of disease mortality rate, but it is necessary to be alert to outliers in heart failure.



## Predictive Modelling





## Model Evaluation

Key findings:






## Conclusion







## References

Public Health Scotland (NHS Scotland, 2024) -- Scottish Heart Disease Statistics -- Heart Disease Mortality by Council Area
https://www.opendata.nhs.scot/dataset/scottish-heart-disease-statistics/resource/fc7b42f1-4de6-48dd-b640-613fb0aa427d
accessed (17/10/2025)

Public Health Scotland (NHS Scotland, 2020) -- Council Area 2011 - Council Area 2019 
https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/967937c4-8d67-4f39-974f-fd58c4acfda5
accessed (7/11/2025)
