---
title: "Project title"
author: "by Artery Party: Roderick Yu, Yufei Xiang, Hoddy Zhu, Lincoln Luk"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, include = FALSE}
library(tidyverse)
library(tidymodels)


```


```{r load-data, include=FALSE}
# load your data here and any pre-processing/cleaning 
#  that you need for your report.

MArea <- read_csv("data/HDM_CouncilArea.csv")
CouncilArea <- read_csv("data/CouncilArea.csv")


Data<- MArea %>%
  select(Year,CA,AgeGroup,Sex,Diagnosis,CrudeRate,EASR) %>%
  drop_na()



```


## Research Question

Describe your research question here. Include why this question might be of interest. 
--------------------------------------------------------------------------------------------------
Our main research question is to explore the overall trends of heart disease mortality in Scotland, and to identify how mortality rates vary across different demographic categories, such as gender, geographical area, years.

Our aim was to find any patterns or if there were more predominant groups of people which show higher mortality rates.





## Data

Briefly describe your dataset here - include a description of the key variables you will be investigating. 
Include a citation for your data. 
See http://libraryguides.vu.edu.au/c.php?g=386501&p=4347840 for guidance on proper citation for datasets. 
If you got your data off the web, make sure to state the retrieval date.
--------------------------------------------------------------------------------------------------
Data Overview

We are using the 'Heart Disease Mortality by Council Area' dataset from the Scottish Heart Disease Statistics from the Public Health Scotland NHS website. 

This dataset consists of 17820 rows and 14 variables (columns).

To support our research aim, we made additional variables to categorise our data better and so we could analyse and evaluate the data more easily.
  - 
  - 


To focus on our main indicators of categories, we mainly used the variables: CA (Council Area code), Sex, Year and AgeGroup.





## Findings

The write-up of your project and findings go here. 

Think of this as the text of your presentation with some extra detail to cover what there was not time to discuss in the presentation. This might include any assumptions you made when doing your analysis, any limitations of the work you have done or any ideas you have for future work. Feel free to split this section into subsections to make it easier to read. 

The length should be roughly 1,500 words. 
If you want to use a word count addin, you can install this by copying and pasting the following into RStudio:

`devtools::install_github("benmarwick/wordcountaddin", type = "source", dependencies = TRUE)`

You will then need to restart RStudio. Once you have done that, select the text you want to count the words of, go to Addins, and select the `Word count` addin.
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 1,500 words, you're good! 
The addin will ignore code chunks and only count the words in prose. If you don't want to use the addin you can always copy and paste the text into Microsoft Word to do a Word Count!

You can also load your data here and present any analysis results / plots.
Make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself, in which case you should show your code.


---------------------------------------------------------------


#Data Visualisations

1. Geographical area

To start off, we tried to find if there were differences in mortality rate for each region of Scotland.
We looked at the dataset CouncilArea from CouncilArea.csv to match with the council area codes in MArea (the original dataset)


2. EASR

The EASR is the European Age-Sex standardised rate per 100,000 population. It is adjusted for the age distribution using a standard population (100,000) and provides a more valid and unbiased comparison than crude rate.


We used the EASR and plotted it against crude rate to see how reliable and it was and we plotted a line of best fit, which had a gradient of 1. This shows that crude rate is very reliable and we can confidently use it to compare evaluate the data, as their correlation is near perfect.


3. Sex 

We 


4. Sex and Age




5. Year




Predictive Modelling




Model Evaluation






Conclusion






## References

Public Health Scotland -- Scottish Heart Disease Statistics -- Heart Disease Mortality by Council Area (NHS)
https://www.opendata.nhs.scot/dataset/scottish-heart-disease-statistics/resource/fc7b42f1-4de6-48dd-b640-613fb0aa427d

Public Health Scotland -- Council Area 2011 - Council Area 2019 
https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/967937c4-8d67-4f39-974f-fd58c4acfda5
